 ---
title: "Adaptation to limitation"
author: "Annegreet Veeken"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
subtitle: "pre-liminary analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	fig.height = 7,
	fig.width = 7
)
options(knitr.kable.NA = '')
```

```{r data-prep}
## Load packages
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(readxl)) install.packages("readxl")
if (!require(TNRS)) install.packages("TNRS")
if (!require(GIFT)) remotes::install_github("https://github.com/BioGeoMacro/GIFT",
                        dependencies = TRUE)
if (!require(rnaturalearth)) install.packages("rnaturalearth")
if (!require(rnaturalearthdata)) install.packages("rnaturalearthdata")
if (!require(sf)) install.packages("sf")
if (!require(dggridR)) install.packages("dggridR")
if (!require(easystats)) install.packages("easystats", dependencies = TRUE)
if (!require(hrbrthemes)) install.packages("hrbrthemes")
if (!require(viridis)) install.packages("viridis")
if (!require(ggeffects)) install.packages("ggeffects")
if (!require(effects)) install.packages("effects")
if (!require(patchwork)) install.packages("patchwork")

## load data
# Globnut species
dir <- "C:/Users/3768651/OneDrive - Universiteit Utrecht/Documents/Data/"
spec <- read.csv(paste0(dir,"Globnut_offline/_GLOBNUT1.0/GlobNut1.0_species.csv")) %>% 
  mutate(species_new = word(species_new, 1,2)) %>% 
  mutate(genus_new = word(species_new, 1)) %>% 
  filter(vascular_plant == 1) %>% 
  filter(!family_new == "") %>% 
  filter(!species_new == "") %>% 
  group_by(family_new, genus_new, species_new) %>% 
  summarise(nobs = n_distinct(plot_ID)) 
fungalroot <- read_xlsx(paste0(dir,
                              "FungalRoot/nph16569-sup-0002-tabless1-s4.xlsx"),
                        sheet = 3, skip = 2)
werner <- read_xlsx(paste0(dir,"Werner_Traits/pnas.1721629115.sd01.xlsx"))
werner <- werner %>% 
  mutate(species_orig = str_replace(species, pattern = "_", 
                               replacement = " ")) %>% 
  dplyr::select(species_orig, parasite, nfix, carnivory, cluster)
nodb <- read_xlsx(paste0(dir, "NodDB/NodDB database v1.3b.xlsx"), 
                  skip = 1, sheet = 1)

## Download and prepare GIFT
# trait_meta <- GIFT_traits_meta()
# gift_raw <- GIFT_traits(trait_IDs = c("1.5.1","4.10.1", "4.5.1"))
# # harmonize nomeclature
# gift <- gift_raw %>% 
#   dplyr::select(species_orig = work_species, parasite = trait_value_1.5.1, carnivory = trait_value_4.10.1, nfix = trait_value_4.5.1)
# saveRDS(gift, "outputs/GIFT_traits.rds")
gift <- readRDS("outputs/GIFT_traits.rds")
```

```{r}
# gift <- readRDS("outputs/GIFT_traits.rds")
# gift_harm <- TNRS(unique(gift$species_orig), sources = c("wfo"))
# saveRDS(gift_harm, "outputs/TNRS/GIFT_harmonization.rds")
gift_harm <- readRDS("outputs/TNRS/GIFT_harmonization.rds")
gift_names <- gift_harm %>% 
  filter(!Accepted_name == "") %>% # no name found, so no taxonomy
  dplyr::select(Name_submitted, species = Accepted_species, 
                family = Accepted_family) %>% 
  mutate(genus = word(species, 1)) %>% 
  filter(!species == ""| is.na(species)) 
gift <- gift %>% 
  left_join(gift_names, by = c("species_orig" = "Name_submitted"))  %>% 
  select(-species_orig) %>% 
  group_by(species) %>% 
  # duplicate arrise due to name harmonization 
  summarise(nfix = paste(na.omit(unique(nfix)), collapse = ", "),
          carnivory = paste(na.omit(unique(carnivory)), collapse = ", "),
          parasite = paste(na.omit(unique(parasite)), collapse = ", ")) %>% 
  # remove duplicate rows with ambiguous trait info
  filter(!nfix == "no, yes") %>% 
  filter(parasite %in% c("independent, obligatory", "independent, obligatory, facultative",  "obligatory, independent", "independent, facultative"))

## Prepare werner
# werner_harm <- TNRS(unique(werner$species_orig), sources = c("wfo"))
# saveRDS(werner_harm, "outputs/TNRS/werner_harmonization.rds")
werner_harm <- readRDS("outputs/TNRS/werner_harmonization.rds")
werner_names <- werner_harm %>% 
  filter(!Accepted_name == "") %>% # no name found, so no taxonomy
  dplyr::select(Name_submitted, species = Accepted_species, 
                family = Accepted_family) %>% 
  mutate(genus = word(species, 1)) %>% 
  filter(!species == ""| is.na(species)) %>% 
  as_tibble()
werner <- werner %>% 
  left_join(werner_names, by = c("species_orig" = "Name_submitted")) %>% 
  select(-species_orig) %>% 
  distinct()

## Prepare Nodb
# nodb_harm <- TNRS(nodb$genus, sources = "wfo")
# saveRDS(nodb_harm,"outputs/TNRS/nodb_harmonization.rds")
nodb_harm <- readRDS("outputs/TNRS/nodb_harmonization.rds")
nodb <- nodb %>% 
  left_join(nodb_harm[,c("Name_submitted", "Accepted_name")], 
            by = c("genus" = "Name_submitted")) %>% 
  # harmonisation merges genera, creating duplicates for consensus column, check if conflicts arise
  group_by(Accepted_name) %>% 
  summarise(nfix = paste(unique(`Consensus estimate`), collapse = ", "),
            n = n_distinct(`Consensus estimate`)) %>%  #used to check for conflicts
  # Zygia has conflict between consensus estimates, chose likely_Rhizobia instead of None, n studies is highest for this
  mutate(nfix = case_match(nfix, "Rhizobia, None"~ "likely_Rhizobia",
                                "Rhizobia, likely_Rhizobia" ~ "likely_Rhizobia",
                                "None, unlikely_Rhizobia" ~ "unlikely_Rhizobia",
                                "likely_present, Present" ~ "likely_present",
                                "likely_Rhizobia, Rhizobia" ~ "likely_Rhizobia",
                                .default = nfix)) %>% 
  dplyr::select(Accepted_name, nfix) %>% 
  filter(!Accepted_name == ""| is.na(Accepted_name)) %>% 
  distinct()

## Prepare Mycorrhiza 
# funroot_harm <- TNRS(fungalroot$Genus, sources = "wfo")
# saveRDS(funroot_harm, "outputs/TNRS/fungalroot_harmonization.rds")
funroot_harm <- readRDS("outputs/TNRS/fungalroot_harmonization.rds")
funroot_names <- funroot_harm %>% 
  filter(!Accepted_name == "") %>% # no name found, so no taxonomy
  dplyr::select(Name_submitted, genus = Accepted_name, 
                family = Accepted_family) %>% 
  filter(!genus == "") %>% 
  as_tibble() 
fungalroot <- fungalroot %>% 
  left_join(funroot_names, by = c("Genus" = "Name_submitted")) %>% 
  distinct()


## Compile to one data set
# check GIFt first then supplement by Werner
cat_traits <- spec %>% 
  left_join(gift, by = c("species_new" = "species")) %>% 
  left_join(werner, by = c("species_new" = "species"), 
            suffix = c("_gift","_werner")) %>% 
  left_join(nodb, by = c("genus_new" = "Accepted_name"), relationship = "many-to-many") %>%   
  left_join(fungalroot, by = c("genus_new" = "Genus"),
                                           relationship = "many-to-many") %>%  
  mutate(species_new = species_new,
         family_new = family_new,
         nobs = nobs,
         carnivory = case_when(carnivory_gift == "non-carnivorous" ~ 0,
                               carnivory_werner == "No" ~  0,
                               carnivory_gift == "carnivorous" ~ 1,
                               carnivory_werner == "Yes" ~ 1,
                               is.na(carnivory_gift) & is.na(carnivory_werner) ~ NA),
         parasite = case_when(parasite_gift == "independent" ~ 0,
                              parasite_gift == "obligatory" | parasite_gift == "facultative" ~ 1,
                              parasite_werner == "No" ~  0,
                              parasite_werner == "Yes" ~ 1,
                              is.na(parasite_gift) & is.na(parasite_werner) ~ NA),
         nfix = case_when(nfix %in%  c("Frankia","likely_Rhizobia","Rhizobia") ~ 1,# check nodb first
                          nfix == "None" ~ 0,
                          nfix_gift == "no" ~ 0, # then gift
                          nfix_gift == "yes" ~ 1,
                          nfix_werner == "No" ~  0, # them werner
                          nfix_werner == "Yes" ~ 1,
                          is.na(parasite_gift) & is.na(parasite_werner) ~ NA),
         cluster = if_else(cluster == "Yes", 1, 0),
         AM = if_else(`Mycorrhizal type` %in% c("AM","NM-AM", "NM-AM, rarely EcM", "EcM-AM"), 1,0),
         OM  = if_else(`Mycorrhizal type` == "OM", 1,0),
         NM  = if_else(`Mycorrhizal type` %in% c("NM", "nM","NM-AM", "NM-AM, rarely EcM"), 1,0),
         EcM  = if_else(`Mycorrhizal type` %in% c("EcM","EcM-AM"), 1,0),
         ErM  = if_else(`Mycorrhizal type` == "ErM", 1,0),
         .keep = "none")

trait_cover <- cat_traits %>% 
  ungroup() %>% 
    summarise(across(c("carnivory","parasite","nfix","AM","OM","NM","EcM","ErM"),
                     ~sum(!is.na(.))/n_distinct(species_new)) %>% round(2))
```


```{r}
npk <- read.csv(paste0(dir,"/Globnut_offline/_GLOBNUT1.0/GlobNut1.0_nutrients.csv"))
meta <- read.csv(paste0(dir,"Globnut_offline/_GLOBNUT1.0/GlobNut1.0_metadata.csv")) 
adapt <- read.csv(paste0(dir,"/Globnut_offline/_GLOBNUT1.0/GlobNut1.0_species.csv")) %>% 
   mutate(species_new = word(species_new, 1,2)) %>% 
  filter(vascular_plant == 1) %>% 
  filter(!family_new == "") %>% 
  filter(!species_new == "") %>% 
  select(plot_ID, species_new) %>% 
  left_join(cat_traits, by = "species_new") %>% 
  left_join(npk %>% 
              select(plot_ID, biomass, NP), by = "plot_ID") %>% 
  left_join(meta %>% 
              select(plot_ID, lon, lat), 
            by = "plot_ID")
```


### Comparison among functional types
```{r}
adapt_long <- adapt %>% 
  select(plot_ID, species_new, NP, biomass,lat,lon, nfix, carnivory,
         parasite, cluster,AM, OM, NM, EcM, ErM) %>%
  pivot_longer(nfix:ErM, names_to = "trait",
               values_to = "value") %>%
  drop_na() %>% 
  filter(value == 1) %>%
  group_by(trait) %>%
  mutate(trait = factor(trait, levels = c("AM","OM","NM","EcM","ErM","nfix",    "parasite","carnivory","cluster"),
                        labels = c("AM" = "AM", "OM" = "OM", "NM" = "NM",
                                    "EcM" = "EcM", "ErM" = "ErM", "nfix" = "N-fixation", 
                                    "parasite" = "Parasitism", "carnivory" = "Carnivorism", 
                                    "cluster" = "Cluster roots")))

adapt_long %>% 
  group_by(trait) %>% 
  summarise(`NP mean` = mean(NP) %>% round(2), 
            `SD NP` = sd(NP) %>% round(2), 
            `Biomass mean` = mean(biomass) %>% round(2), 
            `SD biomass` = sd(biomass) %>% round(2), 
            `N observations` = n_distinct(plot_ID), 
            `N species` = n_distinct(species_new)) %>% 
  knitr::kable()

# N/P
m1 <- aov(log(NP) ~ trait, data = adapt_long)
 
```
The ANOVA (formula: log(NP) ~
trait) suggests that:

  - The main effect of trait is
statistically significant and small
(F(8, 132788) = 209.20, p < .001;
Eta2 = 0.01, 95% CI [0.01, 1.00])

Effect sizes were labelled
following Field's (2013)
recommendations.
```{r results='asis'}
# par(mfrow = c(1,2))  
# hist(m1$residuals, breaks = 100)  
# plot(m1, which = 2)  
m2 <- lm(log(NP) ~ trait, data = adapt_long)

ggplot(adapt_long, aes(y = trait, x = NP, color = trait, fill = trait)) +
    geom_violin(width = 1) +
    geom_boxplot(width = 0.3, color = "grey", alpha = 0.2) +
    scale_fill_viridis(discrete = TRUE) +
    scale_color_viridis(discrete = TRUE) +
    scale_y_discrete("",  limits = rev) +
    scale_x_continuous("N/P") +
    theme_bw() +
    theme(legend.position="none") 
ggeffect(m2) %>% plot() + ylab("N/P (log)") + xlab("") 

# Biomass
m3 <- aov(log(biomass) ~ trait, data = adapt_long)

```
The ANOVA (formula: log(biomass) ~
trait) suggests that:

  - The main effect of trait is
statistically significant and very
small (F(8, 132788) = 119.33, p <
.001; Eta2 = 7.14e-03, 95% CI
[6.35e-03, 1.00])
Effect sizes were labelled
following Field's (2013)
recommendations.

```{r}
# par(mfrow = c(1,2))  
# hist(m3$residuals, breaks = 100)  
# plot(m3, which = 2)  
m4 <- lm(log(biomass) ~ trait, data = adapt_long)

ggplot(adapt_long, aes(y = trait, x = biomass, color = trait, fill = trait)) +
    geom_violin(width = 1) +
    geom_boxplot(width = 0.3, color = "grey", alpha = 0.2) +
    scale_fill_viridis(discrete = TRUE) +
    scale_color_viridis(discrete = TRUE) +
    scale_y_discrete("", limits = rev) +
    scale_x_continuous("Biomass (g/m2)") +
    theme_bw() +
    theme(legend.position="none") 

ggeffect(m4) %>% plot() + ylab("Biomass (g/m2) (log)") + xlab("") 

```


### Spatial distribution
```{r results='asis'}
world <- ne_countries(scale = "medium", returnclass = "sf")
eurasia <- world[world$continent %in% c("Europe", "Asia"),]
dggs <- dgconstruct(res = 8, metric = FALSE, resround = 'down')

trait_geog <- function(trait_name){
sf_coords <- adapt_long %>% 
  filter(trait == trait_name) %>% 
  filter(value == 1) %>% 
  drop_na(lon,lat) 
sf_coords$cell <- dgGEO_to_SEQNUM(dggs,sf_coords$lon,sf_coords$lat)$seqnum
sf_coords <- sf_coords %>% 
  group_by(cell) %>% 
  summarise(count = n_distinct(plot_ID)) 
grid_par  <- dgcellstogrid(dggs,sf_coords$cell)
grid_par <- merge(grid_par,sf_coords,by.x="seqnum", by.y="cell")

p <- ggplot() +
  geom_sf(data = eurasia, fill = "grey90", col = "grey90") +
  geom_sf(data = grid_par, aes(fill = count)) +
  scale_fill_viridis_c("Number of plots") + 
    coord_sf(crs = "+proj=lcc +lat_1=45 +lat_2=55 +lat_0=50 +lon_0=35",
           xlim = c(-4000000, 4500000), ylim = c(-1000000, 4500000)) +
  ggtitle(paste("Distribution of trait in GlobNut -", trait_name)) + 
  theme_bw() 
print(p)
}
map(unique(adapt_long$trait), ~trait_geog(trait_name = .x))

```


### Species list
```{r results='asis'}

spec_list <- function(trait_name){
t <- adapt_long %>% 
  filter(trait == trait_name) %>%
  filter(value == 1) %>% 
  group_by(species_new) %>% 
  summarise(`N observations` = n_distinct(plot_ID),
    `Mean N/P` = mean(NP, na.rm = TRUE) %>% 
      round(digits = 1),
            `SD N/P` = sd(NP, na.rm = TRUE) %>% 
      round(digits = 1),
            `Mean biomass` = mean(biomass, na.rm = TRUE) %>% 
      round(digits = 1),
            `SD biomass` = sd(biomass, na.rm = TRUE) %>% 
      round(digits = 1)
    ) 
knitr::kable(t, caption = paste("Species level N/P and biomass -",trait_name))
}
map(unique(adapt_long$trait), ~spec_list(trait_name = .x))


```
### Presence in habitat types
```{r}
eunis <- read.csv(paste0(dir, "Globnut_offline/_GLOBNUT1.0/Globnut1.0_EUNIS.csv"))
red_list <- read_xlsx("C:/Users/3768651/OneDrive - Universiteit Utrecht/Documents/Data/Habitat_attributes/Red List of habitats enhanced by EEA 2022.xlsx", sheet = "Terrestrial")
eunis <- left_join(eunis %>% 
                     select(plot_ID, habitat, label = eunis_label3),
                   red_list %>% 
                     select(habitat =`EUNIS 2019/2021 code`, 
                            red_list=`Overall category EU28+`),
                   by = "habitat") %>% 
  group_by(habitat) %>% 
  mutate(n = n_distinct(plot_ID)) %>% 
  filter( n > 50)
redlist_adap <- adapt_long %>% 
  left_join(eunis, by = "plot_ID") %>% 
  mutate(red_list = factor(red_list, levels = c("Endangered", "Near Threatened",
                                                "Vulnerable", "Least Concern")))

redlist_adap %>% 
  group_by(habitat, label, red_list, trait) %>% 
  summarise(`N species` = n_distinct(species_new),
            `mean NP` = mean(NP) %>% round(1),
            `sd NP` = sd(NP) %>% round(1)) %>% 
  filter(!is.na(label)) %>%
  knitr::kable()

redlist_adap %>% 
    group_by(habitat, label, red_list, trait) %>% 
  summarise(`N species` = n_distinct(species_new),
            `mean NP` = mean(NP) %>% round(1)) %>% 
  filter(!is.na(red_list)) %>% 
  ggplot(aes(x = red_list, y = `N species`, group = habitat, fill = habitat)) +
     geom_bar(stat="identity", color="black", position="dodge") +
        theme_minimal() +
  xlab("") +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  facet_wrap(~trait, scales = "free_y")
```


### Testing agains null expectation
(not done yet)
```{r eval=FALSE, include=FALSE}
data <- spec_abun %>% 
  drop_na(NP, lat, lon) %>% 
  replace_na(list(carnivory = 0,
                  parasite = 0,
                  nfix = 0,
                  cluster = 0,
                  AM = 0,
                  OM = 0,
                  NM = 0,
                  EcM = 0,
                  ErM = 0)) %>% 
  drop_na(NP) %>% 
  select(plot_ID, lat, lon, NP, carnivory, parasite, nfix, cluster, AM, OM, NM,
         EcM, ErM) %>% 
  distinct() 

# Function for null model
generate_null_np <- function(dat) {
  dat$NP_null <- sample(dat$NP)
  return(dat)
}

# Run null model many times
n_permutations <- 999
null_results <- replicate(n_permutations, {
  null_data <- generate_null_np(data)
  
  # Calculate mean difference for each strategy
  list(
    carnivory = mean(null_data$NP_null[null_data$carnivory == 1]) - 
                  mean(null_data$NP_null[null_data$carnivory == 0]),
    parasite = mean(null_data$NP_null[null_data$parasite == 1]) - 
                 mean(null_data$NP_null[null_data$parasite == 0]),
    nfix = mean(null_data$NP_null[null_data$nfix == 1]) - 
              mean(null_data$NP_null[null_data$nfix == 0]),
    cluster = mean(null_data$NP_null[null_data$cluster == 1]) - 
              mean(null_data$NP_null[null_data$cluster == 0]),  
    AM = mean(null_data$NP_null[null_data$AM == 1]) - 
                   mean(null_data$NP_null[null_data$AM == 0]),
    OM = mean(null_data$NP_null[null_data$OM == 1]) - 
              mean(null_data$NP_null[null_data$OM == 0]),
    NM = mean(null_data$NP_null[null_data$NM == 1]) - 
                   mean(null_data$NP_null[null_data$NM == 0]),
    EcM = mean(null_data$NP_null[null_data$EcM == 1]) - 
                   mean(null_data$NP_null[null_data$EcM == 0]),
    ErM = mean(null_data$NP_null[null_data$ErM == 1])
  )
}, simplify = FALSE)

# Calculate observed differences
observed_diffs <- list(
 carnivory = mean(data$NP_null[data$carnivory == 1]) - 
                  mean(data$NP_null[data$carnivory == 0]),
    parasite = mean(data$NP_null[data$parasite == 1]) - 
                 mean(data$NP_null[data$parasite == 0]),
    nfix = mean(data$NP_null[data$nfix == 1]) - 
              mean(data$NP_null[data$nfix == 0]),
    cluster = mean(data$NP_null[data$cluster == 1]) - 
              mean(data$NP_null[data$cluster == 0]),  
    AM = mean(data$NP_null[data$AM == 1]) - 
                   mean(data$NP_null[data$AM == 0]),
    OM = mean(data$NP_null[data$OM == 1]) - 
              mean(data$NP_null[data$OM == 0]),
    NM = mean(data$NP_null[data$NM == 1]) - 
                   mean(data$NP_null[data$NM == 0]),
    EcM = mean(data$NP_null[data$EcM == 1]) - 
                   mean(data$NP_null[data$EcM == 0]),
    ErM = mean(data$NP_null[data$ErM == 1])
)

# Calculate p-values
calculate_pvalue <- function(null_dist, obs_value) {
  mean(abs(null_dist) >= abs(obs_value))
}

# Combine results
strategies <- unique(adapt$trait)
results <- data.frame(
  strategy = strategies,
  observed_diff = unlist(observed_diffs),
  p_value = sapply(strategies, function(s) {
    null_dist <- sapply(null_results, `[[`, s)
    calculate_pvalue(null_dist, observed_diffs[[s]])
  })
)

```



